---
title: "Introduction to the `ExclusiveLasso` Package"
author:
 - name: Michael Weylandt
   affiliation: Department of Statistics, Rice University
   email: michael.weylandt@rice.edu
date: "2018-10-23"
output: 
  html_document: 
    toc: true
    toc_float: 
      collapsed: false
bibliography: vignettes.bibtex
vignette: >
  %\VignetteIndexEntry{Introduction to the ExclusiveLasso Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE, cache=FALSE}
set.seed(1234)
knitr::opts_chunk$set(cache=TRUE)
```
\[
\DeclareMathOperator{\argmin}{arg\,min}
\DeclareMathOperator{\sign}{sign}
\DeclareMathOperator{\trace}{Trace}
\newcommand{\bI}{\mathbf{I}}
\newcommand{\bM}{\mathbf{M}}
\newcommand{\bo}{\mathbf{o}}
\newcommand{\br}{\mathbf{r}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\bX}{\mathbf{X}}
\newcommand{\by}{\mathbf{y}}
\newcommand{\bz}{\mathbf{z}}
\newcommand{\bw}{\mathbf{w}}
\newcommand{\bW}{\mathbf{W}}
\newcommand{\bg}{\mathbf{g}}
\newcommand{\bV}{\mathbf{V}}
\newcommand{\bmu}{\mathbf{\mu}}
\DeclareMathOperator{\prox}{\textsf{prox}}
\]

The `ExclusiveLasso` package implements the exclusive lasso penalty of Zhou *et
al.* [-@Zhou:2010], which Obozinski and Bach showed is the tightest convex
relaxation of the combinatorial constraint "exactly one non-zero element in each
group" [-@Obozinski:2012, Section 4.2]. For the Gaussian case, we implement
the two algorithms discussed in Campbell and Allen [-@Campbell:2017]:

  - proximal gradient descent, with coordinate descent used to evaluate the
    proximal operator;
  - full coordinate descent algorithm.

The scalings used in the `ExclusiveLasso` package are not exactly those
used by Campbell and Allen, so we describe the algorithms used in the package
in detail below.

## Usage

The `ExclusiveLasso` package implements the exclusive lasso penalty
[@Zhou:2010, @Campbell:2017] for structured variable selection. The interface
and internal design intentionally mimic those of the `glmnet` package [@Friedman:2010]
and, by extension, other sparse regression packages which follow `glmnet`,
notably `ncvreg` for non-convex regularization and `grpreg` for the group
lasso penalty [@Breheny:2011, @Breheny:2015].

We demonstrate its use on a small simulated data set:

```{r}
set.seed(1234)
library(ExclusiveLasso)

n <- 100
p <- 100
g <- 5

groups <- rep(1:g, length.out=p)

Sig <- toeplitz(0.7^((1:p) - 1))
Sig_L <- chol(Sig)

beta <- rep(0, p); beta[1:g] <- runif(g, 2, 3)
X <- matrix(rnorm(n * p), ncol=p) %*% Sig_L
colnames(X) <- paste0(ifelse(beta != 0, "T", "F"), 1:p)
y <- X %*% beta + rnorm(n)

exfit <- exclusive_lasso(X, y, groups=groups)
print(exfit)
```

Here we have significant correlation both within and between groups, but the
correlation is highest between the 5 true positives.

When plotting regularization paths, the last variable to leave the active set
is identified in the legend by default.

```{r fig.width=7, fig.height=7}
plot(exfit)
```

Note that variables from the same group are shown in the same color.

In many cases where the exclusive lasso is used, we have structural knowledge
about the true sparsity pattern and know that we want to select exactly one
variable from each group, so tuning $\lambda$ is not essential. If, however,
we do wish to tune $\lambda$ in a data-driven manner, the `ExclusiveLasso` package
provides a $K$-fold cross-validation function:

```{r fig.width=7, fig.height=7}
exfit_cv <- cv.exclusive_lasso(X, y, groups=groups, parallel=FALSE)
plot(exfit_cv)
```

Running the exclusive lasso is typically quite fast and should not typically
be necessary to run cross-validation in parallel. For large problems or problems
with many groups, it may be necessary to parallize model fits. The `ExclusiveLasso`
package is integrated with the `foreach` package, which provides interfaces
to a number of parallelization schemes.

## Algorithmic Details

We use the "$1/n$"-scaling for penalized regression, as well as including a
factor of $1/2$ in the penalty term:

\[\hat{\beta}_{\text{EL}} = \argmin -\frac{1}{n}\sum_{i=1}^n l(y_i; \bx_i^T\beta) + \lambda\sum_{g \in \mathcal{G}} \frac{\|\beta_g\|_1^2}{2} \tag{EL-GLM} \label{eq:el-glm}\]

where $l(y_i; \bx_i^T\beta)$ is the log-likelihood of the observation $(\bx_i, y_i)$.
For convenience, we will use the script $\ell$ to denote the scaled aggregate *negative*
log-likelihood yielding the problem:^[Below, when observation weights are included,
they will be included in the definition of $\ell$, but not of the $l_i$ terms.]

\[\hat{\beta}_{\text{EL}} = \argmin \ell(\by; \bX\beta) + \lambda\sum_{g \in \mathcal{G}} \frac{\|\beta_g\|_1^2}{2}\]

^[Below, when observation weights are included, they will be included in the definition of $\ell$, but
not of the $l_i$ terms.]

In the Gaussian case, this can be simplified to
\[\hat{\beta}_{\text{EL}} = \argmin \frac{1}{2n}\|\by - \bX\beta\|_2^2 + \underbrace{\lambda \sum_{g \in \mathcal{G}} \frac{\|\beta_g\|_1^2}{2}}_{\lambda * P(\beta)} \tag{EL-Gaussian} \label{eq:el-gaussian}\]

We note that in the case where $\mathcal{G} = \left\{\{1\}, \{2\}, \dots, \{p\}\right\}$
-- that is, there is no meaningful group structure -- the penalty simplifies to the
standard ridge penalty $\frac{\lambda}{2}\|\beta\|_2^2$.

### Proximal Gradient

Campbell and Allen propose the use of a *proximal gradient* algorithm to solve
Problem $\eqref{eq:el-gaussian}$. Proximal gradient algorithms, first proposed in the
sparse regression context by the "ISTA" algorithm of Daubechies *et al.*
[-@Daubechies:2004] and later popularized by Beck and Teboulle [-@Beck:2009],
combine a gradient descent-type update for the smooth part of the objective with
the proximal operator associated with the non-smooth part. They are particularly
common for problems where the non-smooth part has a simple (or even closed form)
solution.^[As will be shown below, the proximal operator of the Exclusive Lasso 
problem is non-trivial and somewhat expensive to evaluate to high precision. For
this reason, an accelerated proximal gradient ("FISTA") algorithm may be implemented
in a future version of the `ExclusiveLasso` package.] Proximal gradient have been
applied to a wide-range of problems. See, *e.g.*,  Parikh and Boyd
[-@Parikh:2014] for a recent review.

For a general penalized regression problem of the form
\[\argmin_{\beta} f(\beta) + \lambda g(\beta)\]
where $f(\cdot)$ is convex and smooth and $g(\cdot)$ is convex, but not smooth,
proximal gradient algorithms work by iterating
\[\beta^{(k)} = \prox_{t_k \lambda g}\left(\beta^{(k-1)} - t_k \nabla f(\beta^{(k-1)}\right)\]
until convergence, where $t_k$ is a step-size which may be fixed or chosen by
a line-search method, and $\prox$ is the proximal operator of $t_k \lambda g$:
\[\prox_{t_k \lambda g}(z) = \argmin_x t_k \lambda g(x) + \frac{1}{2}\|x - z\|_2^2\]

Note that, unlike the use of the proximal operator in ADMM and similar methods
[@Boyd:2011], the step-size $t_k$ appears in the proximal operator.

The name "ISTA" comes from the proximal operator of the $\ell_1$-norm:
\[\prox_{\lambda |\cdot|}(z) = \argmin_x \lambda |x| + \frac{1}{2}(x - z)^2\]
which is the so-called "soft-thresholding" operator:
\[\mathcal{S}_{\lambda}(x) = \begin{cases} x - \lambda & x > \lambda \\ 0 & |x| \leq \lambda \\ x + \lambda & x < -\lambda \end{cases}\]
which gives the iterative soft-thresholding algorithm:
\[\beta^{(k)} = \mathcal{S}_{t\lambda}\left(\beta^{(k-1)} - \frac{t}{n}\bX^T(y - \bX\beta^{(k-1)})\right) = \mathcal{S}_{t\lambda}\left((\bI + t \bX^T\bX/n)\beta^{(k-1)} - t\bX^T\by/n\right)\]
where $t = 1 / \lambda_{\text{max}}(X^TX/n)$ is a constant step-size ensuring convergence.

Hence, for Problem $\eqref{eq:el-gaussian}$, the proximal gradient algorithm becomes
\[\beta^{(k)} = \prox_{t \lambda P(\cdot)}\left((\bI + t\bX^T\bX/n)\beta^{(k-1)} - t\bX^T\by/n\right)\]

Unlike the standard $\ell_1$-penalization case, this proximal operator cannot be
evaluated in closed form and an iterative algorithm must be used to approximate the
proximal operator. This gives rise to a so-called *inexact proximal gradient* scheme,
convergence conditions of which were analyzed by Schmidt *et al.* [-@Schmidt:2011].
Campbell and Allen [-@Campbell:2017] propose the use of a coordinate-descent
scheme to evaluate the proximal operator, described in more detail below.

#### Evaluation of the Proximal Operator

We wish to evaluate the proximal operator:
\[\argmin_{x} \frac{1}{2}\|x - z\|_2^2 + \frac{c}{2}\sum_{g \in \mathcal{G}} \|x_g\|_1^2\]
Note that we fold the constant terms into a single term $c$ in this subsection.

Campbell and Allen [-@Campbell:2017] propose the use of coordinate descent
[@Shi:2016] for this problem and show that it converges, using the sufficient
conditions of Tseng [-@Tseng:2001].

We first note that the proximal operator can be split according to the group structure:[^1]
\[\frac{1}{2}\|x - z\|_2^2 + \frac{c}{2}\sum_{g \in \mathcal{G}} \|x_g\|_1^2 = \sum_{g \in \mathcal{G}} \frac{1}{2}\|x_g - z_g\|_2^2 + \frac{c}{2}\|x_g\|_1^2 \]
so it suffices to derive the coordinate updates for the simpler problem:
\[\frac{1}{2}\|x - z\|_2^2 + \frac{c}{2}\|x\|_1^2\]
Without loss of generality, we derive the update formula for $x_1$:
\[\argmin_{x_1} \frac{1}{2}\|x - z\|_2^2 + \frac{c}{2} \|x\|_1^2 \implies \argmin_{x_1} \frac{1}{2}(x_1 - z_1)^2 + \frac{c}{2}\left(2|x_1| * \|x_{-1}\|_1 + x_1^2\right)\]
Re-grouping and adjusting constant terms, this becomes:
\[\begin{align*}
& \argmin_{x_1} \frac{1}{2}(x_1 - z_1)^2 + \frac{c}{2}\left(2|x_1| * \|x_{-1}\|_1 + x_1^2\right) \\
\implies &\argmin_{x_1} \frac{x_1^2 - 2x_1z_1 + z_1^2 + 2c\|x_{-1}\|_1 * |x_1| + cx_1^2}{2} \\
\implies &\argmin_{x_1} \frac{(1+c)x_1^2 - 2x_1z_1 + z_1^2 + 2c\|x_{-1}\|_1 * |x_1|}{2} \\
\implies &\argmin_{x_1} \frac{1}{2} 2c\|x_{-1}\|*|x_1| + \frac{1}{2}\left[(1+c)x_1^2 - 2x_1z_1\right] \\
\implies &\argmin_{x_1} \frac{c}{1+c}\|x_{-1}\|*|x_1| + \frac{1}{2(1+c)}\left[(1+c)x_1^2 - 2x_1z_1\right] \\
\implies &\argmin_{x_1} \frac{c}{1+c}\|x_{-1}\|*|x_1| + \frac{1}{2}\left[x_1^2 - 2x_1\frac{z_1}{1+c}\right] \\
\implies &\argmin_{x_1} \frac{c}{1+c}\|x_{-1}\|_1 * |x_1| + \frac{1}{2}\left[x_1^2 - 2x_1 \frac{z_1}{1+c} + \frac{z_1^2}{(1+c)^2}\right] \\
\implies &\argmin_{x_1} \frac{c}{1+c} \|x_{-1}\|_1 * |x_1| + \frac{1}{2}\left(x_1 - \frac{1}{1+c}z_1\right)^2
\end{align*}\]
which we recognize as the proximal operator of the scaled absolute value function
\[f(y) = \frac{c}{1+c}\|x_{-1}\|_1 * |y|\] evaluated at $\frac{z_1}{1+c}$
giving the coordinate update:
\[x_i \leftarrow \mathcal{S}_{\frac{c}{1 + c} \|x^{-i}\|}\left(\frac{1}{1 + c} z_i\right)\]

Pulling the common $(1+c)^{-1}$ term out we get:
\[x_i \leftarrow \frac{1}{1+c} \mathcal{S}_{c\|x^{-i}\|}\left(z_i\right)\]

#### Combined Algorithm

Putting these pieces together, we get the following proximal gradient algorithm
for $\eqref{eq:el-gaussian}$.

1. Initialize
    - $\beta^{(0)} = 0$
    - $t = \lambda_{\text{max}}(\bX^T\bX/n)^{-1}$
    - $k = 1$
2. Repeat until convergence:
    - $\bz^{(k-1)} = (\bI + t \bX^T\bX/n)\beta^{(k-1)} - t \bX^T\by/n$
    - For each group $g \in \mathcal{G}$:
        - Initialize $x_g = 0$
        - Repeat until convergence, looping over each $i \in g$:
          \[x_{g, i} = \mathcal{S}_{\frac{2t\lambda}{1 + t\lambda}\|x_{g, -i}\|_1}\left(\frac{t\lambda}{1+t\lambda} z_{g, i}\right)\]
    - $\beta^{(k)} = x$
    - $k := k + 1$.

Generalizing slightly, if we have a vector of offsets $\bo$ and a (diagonal)
weight matrix $\bW$ and an intercept term $\alpha$, the combined algorithm becomes:

1. Initialize
    - $\beta^{(0)} = 0$
    - $t = \lambda_{\text{max}}(\bX^T\bW\bX/n)^{-1}$
    - $k = 1$
2. Repeat until convergence:
    - $\bz^{(k-1)} = (\bI + t \bX^T\bW\bX/n)\beta^{(k-1)} - t \bX^T\bW(\by-\bo)/n$
    - For each group $g \in \mathcal{G}$:
        - Initialize $x_g = 0$
        - Repeat until convergence, looping over each $i \in g$:
          \[x_{g, i} = \mathcal{S}_{\frac{2t\lambda}{1 + t\lambda}\|x_{g, -i}\|_1}\left(\frac{t\lambda}{1+t\lambda} z_{g, i}\right)\]
    - $\beta^{(k)} = x$
    - $\alpha^{(k)} = \left\langle \text{diag}(\bW)/n, \by - \bo - \bX\beta\right\rangle$
    - $k := k + 1$.

### Coordinate Descent

Campbell and Allen [-@Campbell:2017] also propose using a coordinate descent
algorithm to solve $\eqref{eq:el-gaussian}$ directly, similar to that used by
Friedman *et al.* in the `glmnet` package [-@Friedman:2007; -@Friedman:2010]
and by Wu and Lange [-@Wu:2008], among several others, for lasso regression.
As with the proximal operator, this is slightly more complicated because
the penalty term is non-separable, but they show that coordinate descent
converges for the general problem (of which the proximal operator is a special
case), again using the analysis of Tseng [-@Tseng:2001].

Coordinate descent works by sequentially selecting one variable to update and
fixing all others temporarily, minimizing the objective as a function of the
selected variable, and cycling through all variables until convergence. In spite
of its simple structure, many variations of coordinate descent are possible, depending
on how the internal minimization is solved, the strategy by which the active
is chosen, *etc.* Shi *et al.* [-@Shi:2016] review a number of variants. In the
case of sparse regression, significant speed-ups can also be obtained by use of an
"active-set" strategy, where non-zero variables are updated more frequently
than zero variables.

The coordinate updates for the general problem generalize those used
in the proximal operator. Suppose we wish to update $\beta_i$ where $i \in g$.
Then we solve:
\[\argmin_{\beta_i} \frac{1}{2n} \|\by - \bX_{-i}\beta_{-i} - \bx_i\beta_i\|_2^2 + \frac{\lambda}{2}\sum_{g \in \mathcal{G}} \|\beta_g\|_1^2\]
We let $\br = \by - \bX_{-i}\beta_{-i}$ be the "working residual" and omit
penalty terms that do not have $\beta_i$ in them:
\[\argmin_{\beta_i} \frac{1}{2n} \|\br - \bx_i\beta_i\|_2^2 + \frac{\lambda}{2}\left(|\beta_i| + \|\beta_{g,-i}\|_1\right)^2\]
Using similar algebra as before:
\[\begin{align*}
&\argmin_{\beta_i} \frac{1}{2n} \|\br - \bx_i\beta_i\|_2^2 + \frac{\lambda}{2}\left(|\beta_i| + \|\beta_{g,-i}\|_1\right)^2 \\
\implies & \argmin_{\beta_i} \frac{1}{2} \|\br - \bx_i\beta_i\|_2^2 + \frac{n\lambda}{2}\left(|\beta_i| + \|\beta_{g,-i}\|_1\right)^2 \\
\implies & \argmin_{\beta_i} \frac{\|\br\|_2^2 - 2\br^T\bx_i\beta_i + \|\bx_i\|_2^2\beta_i^2 + n\lambda(\beta_i^2 + 2|\beta_i| * \|\beta_{g,-i}\|_1 + \|\beta_{g,-i}\|_1^2}{2} \\
\implies & \argmin_{\beta_i} \frac{\beta_i^2 (\|\bx_i\|_2^2 + n\lambda) - 2\br^T\bx_i\beta_i + 2n\lambda |\beta_i| * \|\beta_{g, -i}\|_1}{2} \\
\implies & \argmin_{\beta_i} n\lambda\|\beta_{g,-i}\|_1 * |\beta_i| +\frac{1}{2}\left(\beta_i^2 (\|\bx_i\|_2^2 + n\lambda) - 2\br^T\bx_i\beta_i \right) \\
\implies & \argmin_{\beta_i} \frac{n\lambda\|\beta_{g,-i}\|_1}{\|\bx_i\|_2^2 + n\lambda} * |\beta_i| +\frac{1}{2(\|\bx_i\|_2^2 + n\lambda)}\left(\beta_i^2 (\|\bx_i\|_2^2 + n\lambda) - 2\br^T\bx_i\beta_i \right) \\
\implies & \argmin_{\beta_i} \frac{n\lambda\|\beta_{g-i}\|_1}{\|\bx_i\|_2^2 + n\lambda} * |\beta_i| +\frac{1}{2}\left(\beta_i^2 - 2\frac{\br^T\bx_i}{(\|\bx_i\|_2^2 + n\lambda)}\beta_i \right) \\
\implies & \argmin_{\beta_i} \frac{n\lambda\|\beta_{g,-i}\|_1}{\|\bx_i\|_2^2 + n\lambda} * |\beta_i| +\frac{1}{2}\left(\beta_i - \frac{\br^T\bx_i}{(\|\bx_i\|_2^2 + n\lambda)} \right)^2 \\
\end{align*}\]
which we recognize as the proximal operator of the scaled absolute value function
\[f(y) = \frac{n\lambda\|\beta_{g-i}\|_1}{\|\bx_i\|_2^2 + n\lambda} * |y|\]
evaluated at $\frac{\br^T\bx_i}{\|\bx_i\|_2^2 + n\lambda}$, which gives the
coordinate update:
\[\beta_i \leftarrow \mathcal{S}_{\frac{n\lambda\|\beta_{g,-i}\|_1}{\|\bx_i\|_2^2 + n\lambda}}\left(\frac{\br^T\bx_i}{\|\bx_i\|_2^2 + n\lambda}\right)\]
Factoring out the common denominator, we get:
\[\beta_i \leftarrow \frac{1}{\|\bx_i\|_2^2 + n\lambda}\mathcal{S}_{n\lambda\|\beta_{g,-i}\|_1}\left(\br^T\bx_i\right)\]

Hence the coordinate descent algorithm is given by:

1. Initialize
    - $\beta = 0$
    - $\br = \by - \bX\beta = \by$
2. Precompute:
    - $\bu$ where $u_i = \bx_i^T\bx_i$
3. Repeat until convergence:
    - For $i \in [p]$:
        - Update working residual: $\br := \br + \bX_i\beta_i$
        - Set $z = \br^T\bx_i$, $\tilde{\lambda} = n\lambda \|\beta_{g,-i}\|_1$
        - Update:
            \[\beta_i = \frac{1}{u_i + n\lambda} \mathcal{S}_{\tilde{\lambda}}(z)\]
        - Update working residual: $\br := \br - \bX_i\beta_i$

In practice, when solving along a grid of values of $\lambda$, convergence can
be much improved by "warm-starting" $\beta^{(0)}$ at the solution for
a nearby value of $\lambda$.

For the general case with offsets $\bo$ and a (diagonal) weight
matrix $\bW$, as well as an unpenalized intercept $\alpha$, the combined
algorithm becomes:

1. Initialize
    - $\beta = 0$
    - $\br = \by - \bo - \bX\beta = \by - \bo$
2. Precompute:
    - $\bu = \text{diag}(\bX^T\bW\bX)$
3. Repeat until convergence:
    - For $i \in [p]$:
        - Update working residual: $\br := \br + \bX_i\beta_i$
        - Set $z = \br^T\bW\bx_i$, $\tilde{\lambda} = n\lambda \|\beta_{g,-i}\|_1$
        - Update:
            \[\beta_i = \frac{1}{u_i + n\lambda} \mathcal{S}_{\tilde{\lambda}}(z)\]
        - Update working residual: $\br := \br - \bX_i\beta_i$
    - Update intercept:
        - Update working residual: $\br: \br + \alpha 1_n$
        - Update intercept: $\alpha := \langle \br, \text{diag}(\bW)/n \rangle$
        - Update working residual: $\br := \br - \alpha 1_{n}$

## Generalized Linear Models with the Exclusive Lasso Penalty

The above methods can be generalized to arbitrary generalized linear models,
though the `ExclusiveLasso` package only currently supports binomial (logistic)
and Poisson GLMs. We implement a proximal gradient method with back-tracking,
similar to that used for Gaussian linear models, which we describe below.

Before we describe the algorithm in detail, we note a useful general result
which covers all the GLMs we consider in this package^[In particular, we restrict
our attention to canonical link functions, natural parameterizations, and
non-overdispersed sampling distributions.] (proof below):

- The gradient of the negative log-likelihood can be written as:
  \[\nabla_{\beta} \ell = -\bX^T (\bW/n) (\by - \bmu) \in \mathbb{R}^{p}\]

where:
- $\bX$ is the design matrix;
- $\bW$ is a diagonal matrix of observation weights; 
- $\by$ is the vector of observed responses; and
- $\bmu$ is the vector of predicted responses. 

As we will see below, $\bmu$ typically has the simple form $g(\bX\beta + \bo)$, 
where $g(\cdot)$ depends only on the GLM family being considered.

To show this, we recall that the general negative log-likelihood for a GLM
can be written as:
\[\ell = \frac{1}{n}\sum_{i=1}^n w_i l(y_i; \bx_i^T\beta + o_i)\]
where $l$ comes from the negative log-likehood of an an exponential family
distribution and hence has the form:
\[l_i = b(\bx_i^T\beta + o_i) - y_i * (\bx_i^T\beta + o_i)\]
Taking derivatives with respect to $\beta_j$, we see that
\[\frac{\partial l_i}{\partial \beta_j} = b'(\bx_i^T \beta + o_i)x_{ij} - y_ix_{ij}\]
Writing this out in a vector form, we see that
\[\nabla_{\beta} l_i = b'(\bx_i\beta + o_i)\bx_i^T - y_i \bx_i^T = \bx_i^T \left[b'(\bx_i^T\beta + o_i) - y_i\right]\]
Hence, since $\ell$ is just a weighted sum of $l_i$, we get
\[\nabla_{\beta} \ell = \frac{1}{n}\sum_{i=1} w_i \nabla_{\beta} l_i = \bX^T\bW\left[b'(\bX\beta + \bo) - \by\right]/n = -\bX^T(\bW/n)(\by - \mu)\]
which shows that $b'$ is just $g$ from above.

We can recover the classical Gaussian model by taking $b(x) = \frac{1}{2}x^2$,
which gives $g(x) = b'(x) = x$. This, in turn, gives the gradient 
\[\nabla_{\beta}\ell = -\bX^T(\bW/n)(\by - \bX^T\beta - \bo)\] which we would 
obtain by taking the gradient of the Gaussian loss directly. Solving the stationary
conditions gives
\[\nabla_{\beta}\ell = \bX^T(\bW/n)\left[\bX\beta + \bo - \by\right] = 0 \implies \beta = (\bX^T\bW\bX)^{-1}\bX^T\bW(\by - \bo)\]
as one would expect.

For logistic regression, we take $b(x) = \log(1 + e^x)$, and so we
recover \[g(x) = b'(x) = \frac{e^x}{1+e^x} = \frac{1}{1+e^{-x}}\] which
maps the real line to the $[0, 1]$ interval.

Finally, for Poisson regression, we take $b(x) = \exp(x)$, and so we recover
\[g(x) = b'(x) = e^x\] which maps the real line to the positive half-line.

Consolidating these results, we see: 

| Family   | Negative Log-Likelihood, $\ell$ | Gradient, $\nabla_{\beta} \ell$ |
|---------:|:--------------------------------|:--------------------------------|
| Gaussian | $\frac{(\by - \bX\beta - \bo)^T\bW(\by - \bX\beta - \bo)}{2n}$ | $\frac{-\bX^T\bW(\by - \bX\beta - \bo)}{n}$ |
| Logistic | $\langle \text{diag}(\bW)/n, \log(1 + \text{exp}(\bX\beta + \bo)) - \by \odot (\bX\beta + \bo)\rangle$ | $-\bX^T\bW(\by - \textsf{inv-logit}(\bX\beta + \bo))/n$
| Poisson  | $\langle \text{diag}(\bW)/n, \exp(\bX\beta + \bo) - \by \odot (\bX\beta + \bo)\rangle$ | $-\bX^T\bW(\by - \text{exp}(\bX\beta + \bo))/n$

where $\odot$ is the elementwise (Hadamard) product of vectors.

### Proximal Gradient Algorithm

The proximal gradient algorithm that we use for GLMs is similar to that for the
Gaussian case, but we introduce a back-tracking step instead of using a fixed
step size. In particular, we use the back-tracking scheme described by Beck and
Teboulle [-@Beck:2010], which also appears in Section 4.2 of Parikh and Boyd
[-@Parikh:2014]. The treatment of the unpenalized intercept term is not as simple
for generalized linear models (since we do not have a closed form for the intercept
in terms of the residuals), so we instead include it as an unpenalized column of
$\bX$ if an intercept appears in the model.

The final algorithm is then:

1. Initialize
    - $\beta^{(0)} = 0$
    - $L = \lambda_{\text{max}}(\bX^T\bW\bX/n)^{-1}$
2. Repeat until convergence:
    - Set: $t = L$
    - Calculate $\nabla \beta^{(k-1)} = -\bX^T(\bW/n)(\by - g(\bX\beta^{(k-1)} + \bo))$
    - Repeat:
        - Let $\bz = \textsf{prox}_{t \lambda}(\beta^{(k-1)} - t \nabla \beta^{(k-1)})$
        - If $\ell(z) \leq \ell(\beta^{(k-1)}) + \langle \nabla \beta^{(k-1)}, z - \beta^{(k-1)} \rangle + \frac{1}{2t} \|z - \beta^{(k-1)}\|_2^2$:
            - $\texttt{break}$
        - Else:
            - Set $t = \alpha * t$ for fixed $\alpha$^[The value of $\alpha$ here
            defaults to $0.8$ and is controlled by the pre-processor macro
            `EXLASSO_BACKTRACK_BETA`.]
    - Set: $\beta^{(k)} = z$, $k: = k + 1$

where $\ell$ is the smooth part of the objective function (*i.e.*, the negative 
log-likelihood), and the proximal operator is evaluated using the coordinate-descent
scheme described above.

### Coordinate Descent Algorithm

In addition to the proximal gradient scheme, we can also employ a (variant of the)
(inexact) proximal Newton method [@Lee:2014]. This method, sometimes also called a
*successive quadratic approximation* approach [@Byrd:2016], works by replacing the
smooth portion of the objective function (the  negative log-likelihood) with a
quadratic approximation and minimizing the approximation, typically with an iterative
algorithm. Once the inner algorithm converges, the approximation is updated and the
process is repeated until convergence. This method has been applied to great effect in
the `glmnet` package for $\ell_1$-penalized GLMs [@Friedman:2010].

Quadratic approximation methods are particularly attractive for generalized
linear models because they reduce to solving a series of penalized
weighted linear regression problems (similar to the IRLS method which solves
unpenalized GLMs by successively fitting weighted least squares) and allow us
to use the efficient algorithm we have already developed for the linear case,
namely coordinate descent. Like `glmnet`, we do not use a back-tracking rule
to guarantee descent.

As with proximal gradient, the implementation of this algorithm is made much
easier by taking advantage of the structure of GLMs. In particular, we note that
the Hessian (second derivative) again has a simple form. Under the same assumptions
as before, we have:

\[\frac{\partial l_i}{\partial \beta_j} = b'(\bx_i^T\beta + \bo_i)x_{ij} - y_ix_{ij}\]

Taking a second derivative with respect to $\beta_k$, we get: 

\[\frac{\partial^2 l_i}{\partial \beta_j \partial \beta_k} = b''(\bx_i^T\beta + \bo_i)x_{ij}x_{ik}\]

This lets us write the general GLM hessian in a particularly simple form: 

\[\mathcal{H}_{\beta} l_i = \bX^T \text{diag} \widetilde{\bW}_i \bX\]

where $\widetilde{\bW}_i$ is a matrix with $b''(\bx_i^T\beta + \bo_i)$ in the $(i, i)$-th
element and zero otherwise. Hence,

\[\mathcal{H}_{\beta} \ell = \sum_{i=1}^n \frac{w_i}{n}\mathcal{H}_{\beta}l_i = \bX^T\left[\bW\widetilde{\bW}/n\right]_i\bX\]

where $\widetilde{\bW} = \sum_{i = 1}^n \widetilde{\bW}_i$, $\bW$ is the diagonal
matrix of observation weights.

For our sequential quadratic approximation / proximal Newton updates, we
replace the smooth portion of the problem with a quadratic approximation, giving: 

\[\argmin_{\beta} \ell(\by, \bX\beta) + \lambda P(\beta) \approx \argmin_{\beta} \ell(\beta_0) + \langle\nabla_{\beta = \beta_0}\ell, \beta - \beta_0\rangle + \frac{(\beta - \beta_0)^T (\mathcal{H}_{\beta = \beta_0}\ell)(\beta - \beta_0)}{2} + \lambda P(\beta)\]

which can be slightly simplified to: 

\[\argmin_{\beta} \ell(\beta_0) + \langle\nabla_{\beta = \beta_0}\ell - \mathcal{H}_{\beta = \beta_0}\beta_0, \beta \rangle + \frac{\beta^T (\mathcal{H}_{\beta = \beta_0}\ell)\beta}{2} + \lambda P(\beta)\]

Using our above expressions for $\nabla \ell$ and $\mathcal{H} \ell$, this becomes

\[\begin{align*}
\argmin_{\beta} \ell(\beta_0) + \langle\nabla_{\beta = \beta_0}\ell - \mathcal{H}_{\beta = \beta_0}\beta_0, \beta \rangle + \frac{\beta^T (\mathcal{H}_{\beta = \beta_0}\ell)\beta}{2} + \lambda P(\beta) \\
\argmin_{\beta} \ell(\beta_0) + \langle -\bX^T\bW(y - g(\bX\beta_0 + \bo))/n - \bX^T\bW\widetilde{\bW}/n\bX\beta_0, \beta\rangle + \frac{\beta^T\bX^T\bW\widetilde{\bW}\bX\beta}{2n} + \lambda P(\beta)\\
\argmin_{\beta} \ell(\beta_0) -\frac{(y - g(\bX\beta_0 + \bo) - \widetilde{\bW}\bX\beta_0)^T\bW\bX\beta}{n} + \frac{\beta^T\bX^T\bW\widetilde{\bW}\bX\beta}{2n} + \lambda P(\beta) \\
\argmin_{\beta} \ell(\beta_0) -\frac{\left(\widetilde{\bW}^{-1}(y - g(\bX\beta_0 + \bo) - \bX\beta_0\right)^T\bW\widetilde{\bW}\bX\beta}{n} + \frac{\beta^T\bX^T\bW\widetilde{\bW}\bX\beta}{2n} + \lambda P(\beta) \\
\end{align*}\]

We note that this quadratic approximation is of the same form as a weighted Gaussian likelihood

\[\argmin_{\beta} \frac{(\by - \bX\beta - \bo)^T\bW(\by - \bX\beta - \bo)}{2n} + \lambda P(\beta) = \argmin_{\beta}\frac{\by^T\bW\by}{2n} - \frac{\by^T\bW\bX\beta}{2n} + \frac{\beta^T\bX^T\bW\bX^T\beta}{n} + \lambda P(\beta)\]

Aligning terms, we write our quadratic approximation as a weighted Gaussian problem

\[\argmin_{\beta} \ell(\beta_0) -\frac{\left(\widetilde{\bW}^{-1}(y - g(\bX\beta_0 + \bo) - \bX\beta_0\right)^T\bW\widetilde{\bW}\bX\beta}{n} + \frac{\beta\bX^T(\bW\widetilde{\bW})\bX\beta}{2n} + \lambda P(\beta) = \argmin_{\beta} \frac{(\bz - \bX\beta - \bo)(\bW\widetilde{\bW}) (\bz - \bX\beta - \bo)}{2n} + \lambda P(\beta)\]

where $\bz = \widetilde{\bW}^{-1}(\by - g(\bX\beta_0 + \bo)) + \bX\beta_0$. We note
that this scheme is very similar to the IRLS (iteratively reweighted least squares)
algorithm used to fit unpenalized GLMs, though IRLS implementations sometimes use
the expected value of $\widetilde{\bW}$ at $\beta_0$ rather than the observed value, 
which gives them a sort of regularized quasi-Newton flavor.

**FIXME** The above drops an offset term somewhere...

Since this problem has been put in Gaussian form, we can re-use the coordinate descent
scheme derived above, yielding our proximal Newton algorithm: 

1. Initialize
    - $\beta = 0$
    - $\eta = \bo$
    - $\mu = g(\eta)$
    - $\widetilde{\bw} = g'(\eta)$
    - $\bz = \widetilde{\bW}^{-1}(y - g(\eta)) + \eta$
    - $\br = \bz - \eta$
2. Repeat until convergence: (outer proximal Newton loop)
    - Precompute: 
        - $\bu = \text{diag}(\bX^T\bW\widetilde{\bW}\bX)$
    - Repeat until convergence: (inner coordinate descent loop)
        - For $i \in [p]$:
            - Update working residual: $\br := \br + \bX_i\beta_i$
            - Set $\zeta = \br^T\bW\widetilde{\bW}\bx_i$, $\tilde{\lambda} = n\lambda \|\beta_{g,-i}\|_1$
            - Update:
                \[\beta_i = \frac{1}{u_i + n\lambda} \mathcal{S}_{\tilde{\lambda}}(\zeta)\]
            - Update working residual: $\br := \br - \bx_i^T\beta_i$
        - Update intercept:
            - Update working residual: $\br := \br + \alpha$
            - Update intercept: $\alpha = \langle \br, \text{diag}(\bW\tilde{\bW})/n \rangle$
            - Update working residual: $\br := \br - \alpha$
    - Update quadratic approximation: 
        - $\eta = \bX\beta + \alpha + \bo$
        - $\mu  = g(\eta)$
        - $\widetilde{\bw} = g'(\eta)$
        - $\bz = \widetilde{\bW}^{-1}(y - g(\eta)) + \eta$
        - $\br = \bz - \eta$

The inverse link $g$ and weighting $g'$ functions are easily calculated, by simple 
extensions of calculations performed above:

| Family   | Inverse Link $g(x)$ | Weight Function $g'(x)$ |
|---------:|:------------------------|:----------------------------|
| Gaussian | $1$                     | $1$                         |
| Logistic | $\textsf{inv-logit}(x) = 1 / (1 + e^{-x})$ | $(1 / (1 + e^{-x})) * (1 - 1 / (1 + e^{-x}))$
| Poisson  | $\text{exp}(x)$ | $\text{exp}(x)$

**FIXME: Add Backtracking for choosing Proximal Newton step size**

## Stationary Conditions

In this section, we derive and re-state some of the stationarity conditions
from Campbell and Allen [-@Campbell:2017] with the scaling conventions used by
the `ExclusiveLasso` package. Many of these are used internally to test the
correctness of the package.

We first derive the subdifferential of the penalty:
\[P(\beta) = \sum_{g \in \mathcal{G}} \frac{\|\beta_g\|_1^2}{2}\]
We abuse notation slightly here and use $x$ and $\{x\}$ interchangeably.

Clearly, this separates group-wise so we first consider the single group case:
\[P(\beta_g) = \frac{\|\beta_g\|_1^2}{2} = \frac{1}{2} \left(\sum_{i \in g } |\beta_{i}|\right)^2\]

If $\beta_{i}$ is non-zero, then $P$ is differentiable with respect to $\beta_i$ and
the subdifferential is given by the classical derivative:
\[\partial_{\beta_i} P(\beta_g) = \frac{1}{2} 2 *\left(|\beta_i| + \|\beta_{-i}\|_1\right) * \sign(\beta_i) = \sign(\beta_i) \|\beta_g\|_1\]

If $\beta_{i}$ is zero, the subdifferential can be evaluted similarly, using
the well-known subdifferential of the absolute value operator at zero:
\[\partial_{\beta_i} P(\beta_g) = \frac{1}{2} * 2 * \left(|\beta_i| + \|\beta_{-i}\|_1\right) * [-1, 1] = [-\|\beta_g\|_1, \|\beta_g\|_1]\]
Since $\beta_i = 0$, $\|\beta_{-i}\|_{1} = \|\beta_g\|_1$ here, which lets
us simplify notation.

Together, this gives
\[\partial_{\beta_i} P(\beta) = \partial(|\beta_i|) * \|\beta_g\|_1 = \begin{cases} \sign(\beta_i) \|\beta_g\|_1 & \beta_i \neq 0 \\ [-\|\beta_g\|_1, \|\beta_g\|_1] &\beta_i = 0 \end{cases}\]

### Proximal Operator

From the above, we can derive KKT conditions for the proximal operator quickly:
\[\bz = \prox_{\lambda P(\cdot)}(\bx) = \argmin_{\bz} \frac{1}{2} \|\bx - \bz\|_2^2 + \frac{\lambda}{2}P(\bz)\]
giving
\[\bz = \prox_{\lambda P(\cdot)}(\bx) \Longleftrightarrow 0 \in (x_i - z_i) + \partial_{z_i}P(\bz)\]

If $z_i \neq 0$, this gives the check:
\[0 = (x_i - z_i) + \lambda \sign(z_i)\|z_g\|_1 \implies x_i = z_i + \lambda \sign(z_i)\|z_g\|_1\]
for $i \in g$. Similarly, if $z_i = 0$, this gives the check:
\[0 \in (x_i - \underbrace{z_i}_{=0}) + \lambda [-\|z_g\|_1, \|z_g\|_1] \implies |x_i| \leq \lambda \|z_g\| \]

A symmetry argument makes clear that if $\bz = \prox_{\lambda P(\cdot)}(\bx)$,
then we must have $\sign(z_i) \in \{0, \sign(x_i)\}$ for all $i$.

We can also derive explicit formulae for the proximal operator in a few cases:

- $\lambda = 0$: $\prox_{0 * P(\cdot)}(\bx) = \bx$
- Groups of size 1: $\prox_{\lambda P(\cdot)}(\bx) = \frac{\bx}{1+\lambda}$

Furthermore, if $\prox_{\lambda P(\cdot)}(\bx)$ has only a single non-zero
element in a group, $z_i \neq 0$, then it must satisfy
\[z_i = \frac{1}{1+\lambda} x_i\]
which generalizes the result above.  This is connected to the result that that
the exclusive lasso reduces to ridge regression when all groups are of size one.

### Gaussian Exclusive Lasso

For the general problem, the KKT conditions are given by:
\[0 \in -\bX^T(\by - \bX\beta)/n + \lambda \partial P(\beta)\]

Restricting attention to the (estimated) support $\hat{\beta}_{\hat{S}}$,
Campbell and Allen [-@Campbell:2017, Proposition 1] show that $\hat{\beta}_{\hat{S}}$ has
a closed-form solution, conditional on the signed support, given by
\[\hat{\beta}_{\hat{S}} = (\bX_{\hat{S}}^T\bX_{\hat{S}}/n + \lambda \bM_\hat{S})^{\dagger}\bX_{\hat{S}}^T\by/n\]
where $\bM_{\hat{S}}$ is a block-diagonal matrix[^2] with blocks given by
$\text{sign}(\hat{\beta}_{\hat{S} \cap g})\text{sign}(\hat{\beta}_{\hat{S} \cap g})^T$.
Note that we have an additional factor of $1/n$ due to the extra $n^{-1}$ in
our problem formulation.

## Statistical Properties

### Degrees of Freedom

Under our scaling, an unbiased estimate of the degrees of freedom is given by
\[\hat{\text{df}} = \trace\left[\bX_{\hat{S}}\left(\bX_{\hat{S}}^T\bX_{\hat{S}} + n * \lambda \bM_{\hat{S}}\right)^{\dagger}\bX_{\hat{S}}^T\right]\] where $\bM$ is as above.

This follows by simple substitution of $\lambda \to \lambda * n$ into Theorem 5
of Campbell and Allen [-@Campbell:2017] to align the two formulations.

To improve numerical stability, we add a small multiple of the identity matrix
to the $\bX_{\hat{X}}^T\bX_{\hat{S}} + n * \lambda \bM_{\hat{S}}$ term before
taking the (pseudo)-inverse.^[ After adding this term, this matrix is always
strictly positive definite so the pseudo-inverse becomes the standard inverse.]

**FIXME** Need to account for observation weights and intercept in DF calculation.

## Benchmarks

We implement a few benchmarks to demonstrate the speed of the coordinate-descent
algorithm used in the `ExclusiveLasso` package.


```{r include=FALSE, echo=FALSE}
set.seed(1234)
```

```{r warning=FALSE, error=FALSE, message=FALSE, results="hide"}
library(microbenchmark)

library(glmnet)
library(ncvreg)
library(grpreg)
library(ExclusiveLasso)

n <- 200
p <- 1000
g <- 10

groups <- rep(1:g, length.out=p)

Sig <- toeplitz(1 + 0.95^(1:p))
Sig_L <- chol(Sig)

X <- matrix(rnorm(n * p), ncol=p) %*% Sig_L
beta <- rep(0, p); beta[1:g] <- runif(g, 2, 3)
y <- X %*% beta + rnorm(n)

MB <- microbenchmark(exlasso=exclusive_lasso(X, y, groups=groups, skip_df=TRUE),
                     glmnet=glmnet(X, y),
                     mcp=ncvreg(X, y, penalty="MCP"),
                     scad=ncvreg(X, y, penalty="SCAD"),
                     grplasso=grpreg(X, y, group=groups),
                     times=20)
```

We skip the degrees of freedom calculation for the exclusive lasso for consistency
with `ncvreg` and `grpreg`. (`glmnet` does return degrees of freedom, but it is
essentially free for `l1` penalized models and incorrect for models with $\ell_2$
penalization.)

This gives the following results on a mid-range Dell XPS 2017 laptop:

```{r echo=FALSE, output='asis', message=FALSE, warning=FALSE, error=FALSE}
library(knitr)

my_table <- function(object, with_respect_to="glmnet"){
    sum <- summary(object)

    unit <- attr(sum, "unit")

    sum_abs <- data.frame(round(sum[, "median"], digits=2))

    colnames(sum_abs) <- paste0("Time (", unit, ")")

    for(i in 1:NCOL(sum)){
        if(is.numeric(sum[1,i])){
            sum[,i] <- sum[,i] / sum[which(sum[,1] == with_respect_to),i]
            sum[,i] <- round(sum[,i], digits=2)
        }
    }

    sum <- sum[,c("expr", "median", "lq", "mean", "uq")]

    sum[,"expr"] <- paste0("**", sum[,"expr"], "**")
    sum[,"median"] <- sprintf("**%.2f**", sum[,"median"])

    colnames(sum) <- c("Method", "Median", "Lower Quartile", "Mean", "Upper Quartile")

    sum <- cbind(sum, sum_abs)

    kable(sum, align=c("l", "r", "r", "r", "r", "r"))
}

my_table(MB)
```

All results are relative to `glmnet`. While the timings are not directly comparable,
since `glmnet`, `ncvreg`, and `exclusive_lasso` all solve different problems, we
see that the performance of all three methods is quite good due to the use of efficient
coordinate descent algorithms. Not surprisingly, the exclusive lasso is slower
than the methods with elementwise penalization (`glmnet`, `mcp`, and `scad`),
but this is not too surprising given the significantly more complex penalty structure.
In practice, the current implementation of the degrees of freedom calculation takes
considerably more time than actual computation of the regularization path.

## References

[^1]: We can also take advantage of the group-structure to evaluate the proximal operator
in parallel, though this is currently not implemented in the `ExclusiveLasso` package.

[^2]: $\bM_{\hat{S}}$ may be a *permuted* block-diagonal matrix if the group structure
does not correspond to adjacent columns of $\bX$.
